---
title: "Sexual Identity and Suicidal Ideation Among U.S. High School Students"
subtitle: "Boston University School of Public Health — Health Equity Hackathon 2026"
author: "Tej Sharma"
date: "2026-02-28"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 6
)
```

---

# Background

This analysis investigates the association between **sexual identity** and **suicidal ideation (SI)**
among U.S. high school students using the **2023 Youth Risk Behavior Survey (YRBS)**.

**Research Question:**
> Is sexual identity associated with suicidal ideation? Does race/ethnicity confound or modify this association?

**Variables:**

| Role | Variable | Type |
|---|---|---|
| Exposure | Sexual Identity | Binary (Heterosexual vs Non-Het/Questioning) |
| Outcome | Suicidal Ideation | Binary (Yes/No) |
| Confounder/EMM | Race/Ethnicity | Categorical (8 groups) |
| Covariates | Age, Sex, Low Grades, Bullying | Binary/Categorical |

---

# Step 1: Load Packages

```{r packages}
library(readr)
library(dplyr)
library(ggplot2)
library(tableone)
library(table1)
```

---

# Step 2: Load Data

```{r load-data}
# Load the YRBS 2023 dataset
# Update this path to where your file is saved
data <- readRDS("/Users/tejsharm/Downloads/yrbs2023_final.rds")

cat("Dataset dimensions:", nrow(data), "rows x", ncol(data), "columns\n")
```

---

# Step 3: Explore Key Variables (Before Cleaning)

```{r explore}
# Look at raw distribution of key variables before any recoding
cat("=== Sexual Identity (raw) ===\n")
table(data$sexual_identity, useNA = "always")

cat("\n=== Suicidal Ideation (raw) ===\n")
table(data$suicideideation_binary, useNA = "always")

cat("\n=== Age (raw) ===\n")
table(data$age, useNA = "always")

cat("\n=== Race/Ethnicity (raw) ===\n")
table(data$raceeth, useNA = "always")
```

---

# Step 4: Data Cleaning

**Decisions made and why:**

- **Age:** Removed students aged 12–13 (codes 1–2) — too young for reliable self-report of sexual identity
- **Sexual Identity:** Collapsed to binary: 0 = Heterosexual, 1 = Non-Heterosexual/Questioning (codes 2–5); code 6 ("Don't Know") set to NA
- **Suicidal Ideation:** Already binary — converted to numeric 0/1
- **Sex:** 0 = Male, 1 = Female
- **Low Grades:** 0 = Good grades (A/B), 1 = Low grades
- **Bullying:** 0 = Not bullied, 1 = Bullied at school

```{r data-cleaning}
table1_clean <- data %>%
  dplyr::select(
    sexual_identity, suicideideation_binary,
    raceeth, age, bulliedschool_binary,
    sex, lowgrades_binary
  ) %>%

  # Remove 12–13 year olds (codes 1 and 2)
  filter(age %in% c("3", "4", "5", "6", "7")) %>%

  mutate(
    # Exposure: Sexual Identity → Binary
    # 0 = Heterosexual, 1 = Non-Het/Questioning, NA = Don't Know
    sexual_identity_cat = case_when(
      sexual_identity == "1"                   ~ 0,
      sexual_identity %in% c("2","3","4","5")  ~ 1,
      TRUE                                     ~ NA_real_
    ),

    # Outcome: Suicidal Ideation → Numeric
    SI = case_when(
      suicideideation_binary == "1" ~ 1,
      suicideideation_binary == "0" ~ 0,
      TRUE                          ~ NA_real_
    ),

    # Covariate: Sex → Numeric (0=Male, 1=Female)
    sex_clean = case_when(
      sex == "1" ~ 1,
      sex == "0" ~ 0,
      TRUE        ~ NA_real_
    ),

    # Covariate: Low Grades → Binary
    lowgrades_clean = case_when(
      lowgrades_binary == "1" ~ 1,
      lowgrades_binary == "0" ~ 0,
      TRUE                    ~ NA_real_
    ),

    # Covariate: Bullied at School → Binary
    bullied_clean = case_when(
      bulliedschool_binary == "1" ~ 1,
      bulliedschool_binary == "0" ~ 0,
      TRUE                        ~ NA_real_
    )
  )

# Verify cleaning
cat("Total students after removing age 12-13:", nrow(table1_clean), "\n\n")

cat("Sexual Identity (cleaned):\n")
table(table1_clean$sexual_identity_cat, useNA = "always")

cat("\nSuicidal Ideation (cleaned):\n")
table(table1_clean$SI, useNA = "always")

cat("\nBullied (cleaned):\n")
table(table1_clean$bullied_clean, useNA = "always")
```

---

# Step 5: Descriptive Statistics (Table 1)

Table 1 describes the characteristics of the study population, both overall and stratified
by sexual identity (the exposure). This helps us see whether the two groups (heterosexual
vs non-heterosexual) differ on key covariates — which tells us which variables we may
need to control for.

```{r table1}
# Remove NAs in exposure variable for stratified table only
# (We only need to know which group each student is in for the stratified table)
yrbs_mod <- table1_clean %>% filter(!is.na(sexual_identity_cat))

# Overall Table 1 — describes the full sample
table1(~ factor(SI) + factor(sex_clean) + factor(age) +
         factor(raceeth) + factor(lowgrades_clean) +
         factor(bullied_clean),
       data = table1_clean)
```

```{r table1-stratified}
# Stratified Table 1 — compares the two exposure groups side by side
# This answers: "Are heterosexual and non-het students different from each other?"
table1(~ factor(SI) + factor(sex_clean) + factor(age) +
         factor(raceeth) + factor(lowgrades_clean) +
         factor(bullied_clean) | sexual_identity_cat,
       data = yrbs_mod)
```

---

# Step 6: SI Prevalence by Sexual Identity Group (All 6 Categories)

Before collapsing to binary, it is useful to visualize SI prevalence across ALL
sexual identity categories to understand the full picture.

```{r plot-prevalence}
prev_data <- table1_clean %>%
  filter(!is.na(sexual_identity), !is.na(SI)) %>%
  mutate(
    si_label = recode(as.character(sexual_identity),
      "1" = "Heterosexual",
      "2" = "Gay/Lesbian",
      "3" = "Bisexual",
      "4" = "Other",
      "5" = "Questioning",
      "6" = "Don't Know"
    )
  ) %>%
  group_by(si_label) %>%
  summarise(
    prevalence = mean(SI, na.rm = TRUE) * 100,
    n          = n(),
    .groups    = "drop"
  )

ggplot(prev_data, aes(x = reorder(si_label, prevalence),
                      y = prevalence,
                      fill = si_label)) +
  geom_bar(stat = "identity", width = 0.6, alpha = 0.85) +
  geom_text(aes(label = paste0(round(prevalence, 1), "%")),
            hjust = -0.2, size = 4, fontface = "bold") +
  coord_flip() +
  scale_fill_brewer(palette = "RdYlBu") +
  scale_y_continuous(limits = c(0, 55)) +
  labs(
    title    = "Prevalence of Suicidal Ideation by Sexual Identity",
    subtitle = "YRBS 2023 | U.S. High School Students (Age 14–18)",
    x        = NULL,
    y        = "Prevalence (%)",
    caption  = "Source: CDC Youth Risk Behavior Survey 2023"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "grey50"),
    plot.caption    = element_text(color = "grey50")
  )
```

---

# Step 7: Association Testing

## 7a. Chi-Square Test

The chi-square test asks: **"Are sexual identity and suicidal ideation statistically associated?"**
It tests whether the distribution of SI differs between heterosexual and non-heterosexual students.

```{r chi-square}
# Cross tabulation with readable labels
cross_tab <- table(table1_clean$sexual_identity_cat, table1_clean$SI)
dimnames(cross_tab) <- list(
  "Sexual Identity"   = c("Heterosexual", "Non-Het/Questioning"),
  "Suicidal Ideation" = c("No", "Yes")
)

# Raw counts
print(cross_tab)

# Row percentages — % with SI in each group
cat("\nRow percentages (% with SI in each group):\n")
round(prop.table(cross_tab, margin = 1) * 100, 1)

# Chi-square test
chisq.test(cross_tab, correct = FALSE)
```

## 7b. Crude (Unadjusted) Odds Ratio

The crude OR tells us the **raw strength of association** between sexual identity and SI,
before accounting for any other variables. It answers: "How much higher are the odds of
SI in non-het students compared to heterosexual students?"

```{r crude-or}
model_crude <- glm(
  SI ~ sexual_identity_cat,
  data   = table1_clean,
  family = binomial(link = "logit")
)

summary(model_crude)

cat("\n=== Crude OR with 95% CI ===\n")
round(exp(cbind(OR = coef(model_crude),
                confint.default(model_crude))), 2)
```

---

# Step 8: Adjusted Logistic Regression

**Why adjust?**
The crude OR may be distorted by other variables (confounders). For example, non-het
students may be more likely to get bullied, and bullying independently increases SI risk.
By adjusting, we isolate the **true** effect of sexual identity.

**Why `relevel()` for raceeth and age only?**
These are the only variables with 3+ categories. Binary variables (sex, grades, bullying)
automatically use 0 as the reference — no manual specification needed.

```{r adjusted-model}
model_adjusted <- glm(
  SI ~ sexual_identity_cat +
       relevel(factor(raceeth), ref = "5") +   # White as reference
       relevel(factor(age),     ref = "5") +   # Age 16 as reference (middle)
       sex_clean +
       lowgrades_clean +
       bullied_clean,
  data   = table1_clean,
  family = binomial(link = "logit")
)

summary(model_adjusted)

cat("\n=== Adjusted OR with 95% CI ===\n")
round(exp(cbind(OR = coef(model_adjusted),
                confint.default(model_adjusted))), 2)
```

---

# Step 9: Confounding Assessment

**Rule:** If the OR changes by more than 10% after adjustment → confounding is present.

```{r confounding-check}
crude_or    <- exp(coef(model_crude)["sexual_identity_cat"])
adjusted_or <- exp(coef(model_adjusted)["sexual_identity_cat"])
pct_change  <- abs(crude_or - adjusted_or) / crude_or * 100

cat("==========================================\n")
cat("         CONFOUNDING ASSESSMENT\n")
cat("==========================================\n")
cat("Crude OR:    ", round(crude_or, 2), "\n")
cat("Adjusted OR: ", round(adjusted_or, 2), "\n")
cat("% Change:    ", round(pct_change, 1), "%\n")
cat("Confounding present (>10%)?",
    ifelse(pct_change > 10, "YES → Report Adjusted OR",
                            "NO  → Crude OR is sufficient"), "\n")
cat("==========================================\n")
```

---

# Step 10: Visualize Crude vs Adjusted OR (Forest Plot)

```{r forest-plot-compare}
or_data <- data.frame(
  Model = c("Crude OR", "Adjusted OR"),
  OR    = c(crude_or, adjusted_or),
  LCI   = c(
    exp(confint.default(model_crude)["sexual_identity_cat", 1]),
    exp(confint.default(model_adjusted)["sexual_identity_cat", 1])
  ),
  UCI   = c(
    exp(confint.default(model_crude)["sexual_identity_cat", 2]),
    exp(confint.default(model_adjusted)["sexual_identity_cat", 2])
  )
)

ggplot(or_data, aes(x = OR, y = Model, color = Model)) +
  geom_point(size = 5) +
  geom_errorbarh(aes(xmin = LCI, xmax = UCI),
                 height = 0.15, linewidth = 1.2) +
  geom_vline(xintercept = 1, linetype = "dashed",
             color = "black", linewidth = 0.8) +
  geom_text(aes(label = paste0("OR = ", round(OR, 2),
                               "\n(", round(LCI, 2), "–", round(UCI, 2), ")")),
            vjust = -1.2, size = 3.8, fontface = "bold") +
  scale_color_manual(values = c("Crude OR"    = "#E74C3C",
                                "Adjusted OR" = "#2980B9")) +
  scale_x_continuous(limits = c(0.5, 6)) +
  labs(
    title    = "Crude vs Adjusted Odds Ratio",
    subtitle = "Association between Sexual Identity and Suicidal Ideation",
    x        = "Odds Ratio (95% CI)",
    y        = NULL,
    caption  = "Adjusted for race/ethnicity, age, sex, low grades, bullying"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "none",
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "grey50"),
    plot.caption    = element_text(color = "grey50")
  )
```

---

# Step 11: Effect Modification Testing (Race/Ethnicity)

**Why test for effect modification?**
We want to know if the association between sexual identity and SI is **the same across
all racial groups**, or if it **differs** by race. This is tested with an interaction term.

**How to read the output:**
Look at lines starting with `sexual_identity_cat:factor(raceeth)`.
If ANY of these p-values < 0.05 → effect modification is present → run stratified models.

```{r interaction-test}
interaction_model <- glm(
  SI ~ sexual_identity_cat * factor(raceeth) +
       relevel(factor(age), ref = "5") +
       sex_clean +
       lowgrades_clean +
       bullied_clean,
  data   = table1_clean,
  family = binomial(link = "logit")
)

# Extract ONLY the interaction terms to make it easier to read
interaction_summary <- summary(interaction_model)$coefficients
interaction_rows <- grep("sexual_identity_cat:factor", rownames(interaction_summary))

cat("=== INTERACTION TERMS (Effect Modification Test) ===\n")
cat("Look for p < 0.05 in the last column\n\n")
round(interaction_summary[interaction_rows, ], 3)

cat("\n=== DECISION ===\n")
p_values <- interaction_summary[interaction_rows, 4]
if (any(p_values < 0.05)) {
  cat("Interaction p < 0.05 FOUND → Effect modification present → Run stratified models\n")
} else {
  cat("All interaction p > 0.05 → NO effect modification\n")
  cat("Race/ethnicity is a CONFOUNDER only\n")
  cat("Report ONE adjusted OR for all groups\n")
}
```

---

# Step 12: Publication-Ready Summary Forest Plot

```{r forest-plot-final}
all_results <- data.frame(
  Variable = c(
    "Sexual Identity\n(Non-Het vs Heterosexual)",
    "Sex\n(Female vs Male)",
    "Low Grades\n(Yes vs No)",
    "Bullying\n(Yes vs No)"
  ),
  OR  = as.numeric(exp(coef(model_adjusted)[
          c("sexual_identity_cat", "sex_clean",
            "lowgrades_clean",     "bullied_clean")])),
  LCI = as.numeric(exp(confint.default(model_adjusted)[
          c("sexual_identity_cat", "sex_clean",
            "lowgrades_clean",     "bullied_clean"), 1])),
  UCI = as.numeric(exp(confint.default(model_adjusted)[
          c("sexual_identity_cat", "sex_clean",
            "lowgrades_clean",     "bullied_clean"), 2]))
)

ggplot(all_results, aes(x = OR, y = reorder(Variable, OR))) +
  geom_point(size = 4, color = "#2C3E50") +
  geom_errorbarh(aes(xmin = LCI, xmax = UCI),
                 height = 0.25, linewidth = 1,
                 color  = "#2C3E50") +
  geom_vline(xintercept = 1, linetype = "dashed",
             color = "red", linewidth = 0.8) +
  geom_text(aes(label = paste0("OR=", round(OR, 2),
                               " (", round(LCI, 2), "–", round(UCI, 2), ")")),
            vjust = -1, size = 3.5, color = "#2C3E50") +
  scale_x_continuous(limits = c(0.5, 5.5)) +
  labs(
    title    = "Adjusted Odds Ratios for Suicidal Ideation",
    subtitle = "YRBS 2023 | U.S. High School Students (Age 14–18)",
    x        = "Odds Ratio (95% CI)",
    y        = NULL,
    caption  = "Adjusted for race/ethnicity and age\nReference: Heterosexual, Male, Good Grades, Not Bullied"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "grey50"),
    plot.caption  = element_text(color = "grey50"),
    panel.grid.minor = element_blank()
  )
```

---

# Step 13: Heatmap — SI Prevalence by Race and Sexual Identity

This plot visualizes the SI prevalence across all combinations of race/ethnicity and
sexual identity, making it easy to spot which groups are most at risk.

```{r heatmap}
heat_data <- table1_clean %>%
  filter(!is.na(SI), !is.na(sexual_identity_cat), !is.na(raceeth)) %>%
  mutate(
    si_label = ifelse(sexual_identity_cat == 0,
                      "Heterosexual", "Non-Het/Questioning"),
    race_label = recode(as.character(raceeth),
      "1" = "Am. Indian/AK Native",
      "2" = "Asian",
      "3" = "Black/African American",
      "4" = "Native Hawaiian/PI",
      "5" = "White",
      "6" = "Hispanic/Latino",
      "7" = "Multiple-Hispanic",
      "8" = "Multiple-Non-Hispanic"
    )
  ) %>%
  group_by(race_label, si_label) %>%
  summarise(prevalence = mean(SI, na.rm = TRUE) * 100, .groups = "drop")

ggplot(heat_data, aes(x = si_label, y = race_label, fill = prevalence)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = paste0(round(prevalence, 1), "%")),
            size = 4, fontface = "bold") +
  scale_fill_gradient(
    low  = "#FFF7BC",
    high = "#B2182B",
    name = "SI Prevalence (%)"
  ) +
  labs(
    title    = "Suicidal Ideation Prevalence by Sexual Identity and Race/Ethnicity",
    subtitle = "YRBS 2023",
    x        = NULL,
    y        = NULL,
    caption  = "Source: CDC Youth Risk Behavior Survey 2023"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey50"),
    axis.text     = element_text(size = 10),
    legend.position = "right"
  )
```

---

# Step 14: Final Results Summary

```{r final-summary}
crude_ci <- exp(confint.default(model_crude)["sexual_identity_cat",])
adj_ci   <- exp(confint.default(model_adjusted)["sexual_identity_cat",])

tab <- prop.table(table(table1_clean$sexual_identity_cat,
                        table1_clean$SI), 1) * 100

cat("══════════════════════════════════════════════\n")
cat("            FINAL RESULTS SUMMARY\n")
cat("══════════════════════════════════════════════\n")
cat("Sample size:                  ", nrow(table1_clean), "\n")
cat("Complete cases (key vars):    ",
    sum(!is.na(table1_clean$SI) &
        !is.na(table1_clean$sexual_identity_cat)), "\n\n")

cat("── Prevalence of Suicidal Ideation ──\n")
cat("Heterosexual students:        ", round(tab[1,2], 1), "%\n")
cat("Non-Het/Questioning students: ", round(tab[2,2], 1), "%\n\n")

cat("── Association Results ──\n")
cat("Crude OR:    ", round(crude_or, 2),
    " (95% CI:", round(crude_ci[1],2), "–", round(crude_ci[2],2), ")\n")
cat("Adjusted OR: ", round(adjusted_or, 2),
    " (95% CI:", round(adj_ci[1],2), "–", round(adj_ci[2],2), ")\n\n")

cat("── Confounding ──\n")
cat("% Change crude→adjusted:      ", round(pct_change, 1), "%\n")
cat("Confounding present:           YES (>10% change)\n\n")

cat("── Effect Modification ──\n")
cat("Interaction with race/ethnicity: NOT present (all p > 0.05)\n")
cat("══════════════════════════════════════════════\n")
```

---

# Interpretation and Conclusion

## Interpretation

**Prevalence:** Non-heterosexual and questioning students showed approximately double
the prevalence of suicidal ideation compared to heterosexual students.

**Crude OR = 4.39:** Before adjusting for any other factors, non-heterosexual/questioning
students had 4.39 times higher odds of suicidal ideation compared to heterosexual students.

**Adjusted OR = 3.35:** After controlling for race/ethnicity, age, sex, low grades, and
bullying, non-heterosexual/questioning students still had 3.35 times higher odds of
suicidal ideation — and the association remained statistically significant.

**Confounding (23.5% change):** The OR decreased from 4.39 to 3.35 after adjustment,
indicating that age, sex, race, grades, and bullying were partially inflating the crude
association. Confounding was present and the adjusted OR should be reported.

**Effect Modification:** No significant interaction between sexual identity and
race/ethnicity was detected (all p-interaction > 0.05). This means the association
between sexual identity and suicidal ideation is consistent across all racial/ethnic groups.

## Conclusion

Non-heterosexual and questioning youth face significantly higher odds of suicidal ideation
compared to their heterosexual peers (Adjusted OR = 3.35, 95% CI: X.XX–X.XX), even after
accounting for demographic and academic factors. This association is consistent across all
racial/ethnic groups. These findings underscore the urgent need for targeted, culturally
sensitive mental health interventions for LGBTQ+ youth in school settings.

## Limitations and Potential Biases

**Selection Bias:** The YRBS is a school-based survey. Students who are absent, homeless,
or have dropped out are not captured — these groups may have even higher rates of SI,
meaning our results may **underestimate** the true prevalence.

**Information Bias:** All data are self-reported. Social desirability bias may lead
students to **underreport** their sexual identity or suicidal ideation, particularly
in less accepting school environments.

**Misclassification Bias:** Sexual identity is measured at a single time point and may
not capture its fluidity in adolescence. This likely causes **non-differential
misclassification**, which biases the OR toward the null (underestimates the true effect).

**Residual Confounding:** Variables not available in this dataset (e.g., family
acceptance, mental health history, neighborhood resources) could still be confounding
the association.

---

*Analysis conducted as part of the Boston University School of Public Health Health Equity Hackathon, February 2026.*
